---
title: "bacteria"
author: "CARLENE IRIMBERE"
date: "2024-10-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A. INTRODUCTION

### BUSSINESS TASK
Thsi work is done for learning purposes. the objectify is to do a exploratory analysis of the data set and record interesting fact discovered. 

## B.PACKAGES REQUIRED FOR THIS PROJECT

```{r}
library(tidyverse)
library(ggthemes)
library(sqldf)
library(rvest)
library(naniar)
library(visdat)
library(arrow)
library(duckdb)



## also installing the dependencies ‘gargle’, ‘ids’, ‘dbplyr’, ‘dtplyr’, ‘forcats’, ‘googledrive’, ‘googlesheets4’, ‘haven’, ‘lubridate’, ‘modelr’, ‘reprex’, ‘rvest’

```


## C. SET WORKING DIRECTORY
```{r}
setwd("I:/My Drive/DATA ANALYTICS/PROJECT PORTFOLIO/HEALTH CARE/bacteria")


getwd()
```

## D. DATASET REQUIRED


#### D.1.Description of the dataset. 

 About Data set
 =============
Data set Overview: 
This data set provides a comprehensive overview of 200 unique bacterial species, highlighting their scientific classification, natural habitats, and potential impacts on human health. 
Data Science Applications


credit: Thid data set was downoaled from KAGGLE



#### Loading datasets


```{r}
bacteria2 <- read_csv("I:/My Drive/DATA ANALYTICS/PROJECT PORTFOLIO/HEALTH CARE/bacteria/bacteria_list_200_original_file.csv")
```



# E. EXPLORATORY ANALYSIS


### E.1 PRE-CLEANING ACTIVITIES



#### 1.1 INSPECTION OF THE DATASET

```{r}
View(bacteria)
```

result: 
upon opening the dataset, I realized one thing, the column where found which is the natural habit for the bacteries, has multiple information and on criteria for achieving tidy data is to have 1 value per record. 

the cleaning I wanted to do, I could not implement it in r studio or sql so I did it in excel since the data is small. for the sake of replication this is how I did it. 

1. created table for all the data
2. using Power query, split the column " where found" with delimiter "and", "comma", "&", in 5 column untill each record had 1 value. 
3. then copied all the under the original "wher found" column making sure to keep the other field intact. 
the number of records went from 200 to 280.
to guarantee that the number of the record is accurate, used the function "counta" to make sure all the values have been well replicated. 

4. the column "where found" was renamed "Natural_habitat" and "Harmful toHumans" to "Harmful_to_Humans"

5. Triming the space and capitolizing each letter of the "natuaral_habitat" field since r studio is case sensitive. 





F. LIMITATION OF THE DATA SET
=============================
in some fields, the value said "intestinal", "genitarinary track" without specifying if it's for humans of aminals.


G. LOADONG CLEANED VERSION OF THE DATA SET 
===========================================

```{r}
BACTERIA <- read_csv("I:/My Drive/DATA ANALYTICS/PROJECT PORTFOLIO/HEALTH CARE/bacteria/bacteria_list_cleaned_version.csv")
```


```{r}
View(BACTERIA)
```


H. PRERIMINARY QUESTIONS
======================

1. Grouping by:
a. the location count to the distribution of the natural_habitat. what is the top 10 of the most found natural habitat?

b. by family, what is the most common family found in the dataset? show the top 10. and in total how many families do we have?

2, what is the ration in term of difference between tha harmful and the non-hamrful to humans?

3. what is the distribution of harmful bacteria across the 10 most common families and if possible to answer this question: the least common families are they mostly harmful or non-harmful to human?

4. and if possible to answer this question: the least common families are they mostly harmful or non-harmful to human? so that we don't neglect them being like, they are the least common so no need to worry about them

5. where the top 10 of harmful bacteria are found and top 10 of the non-harmful bacteria are found?


#### QUESTION 1 
 --------------


1. Grouping by:
a. the location count to the distribution of the natural_habitat. what is the top 10 of the most found natural habitat?

b. by family, what is the most common family found in the dataset? show the top 10.

  A.
 ----

```{r}
BACTERIA |> 
  group_by(Natural_habitat) |> 
  count() |> 
  arrange( desc(n)) |> 
   head( 10) |> 
  mutate( percentage = round (n/nrow( BACTERIA)* 100, 2))
```
show the percentage of it. 

```{r}
BACTERIA |> 
  
```



 VISUALIZATION
 =============
```{r}
BACTERIA |> 
  group_by(Natural_habitat) |> 
  count() |> 
  arrange( desc(n)) |> 
   head( 10) |> 
  ggplot(aes(x= n , y = fct_infreq (Natural_habitat), fill= Natural_habitat)) +
  geom_bar( stat = "identity")
```


for example this won't work if you don't add "identity"

```{r}
BACTERIA |> 
  group_by(Natural_habitat) |> 
  count() |> 
  arrange( desc(n)) |> 
   head( 10) |> 
  ggplot(aes(x= Natural_habitat)) +
  geom_bar( )
```



  B.
 ----
WHAT IS THE TOTAL DISTINCT COUNT OF FAMILIES 
 ==================================
```{r}
BACTERIA |> 
  distinct(Family) |> 
  count()
```
 THE TOTAL COUNT OF THE FAMILIES
  =============================
  
```{r}
  nrow(BACTERIA)
```

 
 
b. by family, what is the most common family found in the data set? show the top 10.

```{r}
BACTERIA |> 
  group_by(Family) |> 
  count() |> 
  arrange( desc(n)) |> 
   head( 10)
```

 VISUALIZATION
 -------------
```{r}
BACTERIA |> 
  group_by(Family) |> 
  count() |> 
  arrange( desc(n)) |> 
   head( 10) |> 
  ggplot(aes(x = n , y = Family, fill= Family)) +
  geom_bar(stat = "identity")
```
 


### QUESTION 2
 -------------


2, what is the ration in term of difference between the harmful and the non-hamrful to humans?

```{r}
BACTERIA |> 
  
  group_by(Harmful_to_Humans) |> 
  count() |> 
  mutate(percentage = round (n/nrow(BACTERIA) * 100, 2)) |> 
  arrange(desc(percentage))
```

 VISUALIZATION 
 -------------
```{r}
BACTERIA |> 
  ggplot(aes( x= fct_infreq( Harmful_to_Humans), fill= Harmful_to_Humans)) +
  geom_bar()
```


### QUESTION 3 
 ==============

3. what is the distribution of harmful bacteria across the 10 most common families and if possible to answer this question: the least common families are they mostly harmful or non-harmful to human?

 COUNT OF THE FAMILIES
 ====================
 a. count of families sorted by the most found
 ---------------------------------------------

```{r}
BACTERIA |> 
  group_by(Family) |> 
  count() |> 
  arrange(desc(n))
```



 b. top 10 out of all those families
 -------------------------------------
select TOP 10 Family,  count (family) count_family
from BACTERIA
group by Family
order by count_family desc

```{r}
BACTERIA |> 
  group_by(Family) |> 
  count() |> 
  arrange(desc(n)) |> 
  head(10)
```

   IMPORTANT NOTE
 ==================
IT WOULD BEST, AFTER ANALYSING THE RESULT OF THESE 2 (ABOVE) QUEUIRES, TO GO WITH TOP 15 SO THAT ALL THE BACTERIA WITH AT LEAST A PREVALENCE SUPERIOR TO 5 ARE REPRESENTED. 

FOR THE SAKE OF THE ANALYSIS, BACTERIA WITH A PREVALENCE ABOVE OR EQUAL TO 5 ARE GOING TO BE CONSIDERED COMMOM OR IN THE TOP 15, THE PREVALENCE BELOW OR EQUAL TO 4 IS GOING TO BE LABELED NON- COMMON. 

 c. top 15 out of all those families
 -------------------------------------

```{r}
BACTERIA |> 
  group_by(Family) |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15)
```


From the top 15 most found bacteria IN COUNT, what is the distribution of harmful to humans bacteria
    or
among the 15 family what is the prevalence of the harmful to human bacteria and non harmful to human bacteria


   harmful
  ----------
```{r}
BACTERIA |> 
  group_by(Family) |> 
  filter(Harmful_to_Humans == "Yes") |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15)
```


 VISUALIZATION 
 ------------
```{r}
BACTERIA |> 
  group_by(Family) |> 
  filter(Harmful_to_Humans == "Yes") |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15) |> 
  ggplot(aes(x = n , y = Family, fill= Family)) +
  geom_bar(stat = "identity")
```


OR: THIS ALSO GIVES THE SAME RESULT. 

```{r}
BACTERIA |> 
   filter(Harmful_to_Humans == "Yes") |> 
  group_by(Family) |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15)
```


 non- harmful
 --------------
```{r}
BACTERIA |> 
   filter(Harmful_to_Humans == "No") |> 
  group_by(Family) |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15)
```
 VISUALIZATION 
 =============
```{r}
BACTERIA |> 
  group_by(Family) |> 
  filter(Harmful_to_Humans == "No") |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15) |> 
  ggplot(aes(x = n , y = Family, fill= Family)) +
  geom_bar(stat = "identity")
```
 
 
 i would like to have a visualization that shows on stacked bar, the top 15 families in bar and the color is Yes or No to being harmful to humans. 
 
### QUESTION 4
 ==============
 
 and if possible to answer this question: the least common families are they mostly harmful or non-harmful to human? so that we don't neglect them being like, they are the least common so no need to worry about them


as we have established in the important note: the least common are going to be the one with a prevalence that is below 4

  STEPS 
 =======
 a. count the families and sorting them in the ascending order and limit to where the count is equal or below 4 


```{r}
BACTERIA |> 
  group_by( Family) |> 
  count() |> 
  filter(n <= 4) |> 
  arrange (desc(n))
```

 b. find the revalence of harmful bacteria in those we selected.
 ----------------------------------------------------------------------


```{r}
BACTERIA |> 
  filter(Harmful_to_Humans == 'Yes') |> 
  group_by( Family) |> 
  count() |> 
  arrange ( n)
```
 
the harmullness of the bacteria in the bracket prevalence equal to 4 
 ----------------------------------------------------------
```{r}
BACTERIA |> 
  filter(Harmful_to_Humans == 'Yes') |> 
  group_by( Family) |> 
  count() |> 
  filter(n == 4) |> 
  arrange ( n)
```
 
 
 
 
 
   c. find the relavence of non-harmful bacteria in those we selected.
  --------------------------------------------------------------
 
```{r}
BACTERIA |> 
  filter(Harmful_to_Humans == 'No') |> 
  group_by( Family) |> 
  count() |> 
  filter(n <= 4) |> 
  arrange ( n)
```
### QUESTION 5
 =============

 5. where the top 15 of harmful bacteria are found and top 15 of the non-harmful bacteria  are found?

   STEPS
 ========

 a. first we find out where the top 15 families live
 -------------------------------------------------------

```{r}
BACTERIA |>
  group_by(Family, Natural_habitat) |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15)
```

  harmful
 ---------

```{r}
BACTERIA |>
  group_by(Family, Natural_habitat) |> 
  filter(Harmful_to_Humans == "Yes") |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15)
```

 non- harmful
 -----------------

```{r}
BACTERIA |>
  group_by(Family, Natural_habitat) |> 
  filter(Harmful_to_Humans == "No") |> 
  count() |> 
  arrange(desc(n)) |> 
  head(15)
```
 

## CONCLUSION ON OUR FINDINGS.
 =============================
 
 After the exploratory analysis, I found out that: 
 
 
  KEYS FINDINGS
  ============
1. In the distribution of the natural habitat,  
Soil comes way far ahead, followed by water, Intestinal Tract, Plants. 
2. Harmful bacteria to humans counts for 50.71% fo the total count of bacteria. whist non-harmful bacteria are	49.29	%

3. The top 3 families most commonly found overall are Enterobacteriaceae	with a leading count of 33 , Streptococcaceae	23			
Flavobacteriaceae	9	

and yet the most commonly found in the harmful category are
Streptococcaceae	19			
Enterobacteriaceae	15			
Flavobacteriaceae	8

and in the non-harmful category
Enterobacteriaceae	18		
Bacillaceae	7		
Acetobacteraceae	5

4, the interesting fact about the less common bacteria, meaning a prevalence below or equal to 4, have an almost 50-50 distribution with the bacteria in the bracket of prevalence equal to 4 being all harmful to human. 

5. whent it comes to the habitat of the bacteria in relationship to their harmulness: 
the top 15 of the harmfull category are mainly found in the  Intestinal Tract,	Genitourinary Tracts, Intestinal and in the Soil.
the top 15 of the non-harmful to human are found mainly in the Intestinal Tract, Soil and in the human gut. 